---
title: "Exploratory Data Analysis"
subtitle: "using R Programming"
author: "Allan R. Lao, MBA, DIT"
date: "Department of ICT </br> `r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["rutgers", "rutgers-fonts"]
    nature:
     
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center") 
```

---


```{r include=TRUE, echo=FALSE}
library(hflights)
library(tidyverse)
df <- as.data.frame(hflights)
```
# The hflights Dataset
Includes data on all flights that departed **Houston, TX in 2011**

```{r}
df <- as.data.frame(hflights)

str(df)
```

>There are 227496 observations; 21 variables
---
# The hflights Dataset
```{r}
knitr::kable(
    head(df, 10),format = 'html'
)
```



#Converting to Factors

**Factor** variables are **categorical** variables that can be either numeric or string variables

```{r}
df$Origin[1:20]
class(df$Origin)
unique(df$Origin)
```

Since Origin appears to have categorical values of **IAH** and **HOU** only then convert to Factor
Examine also DayOfWeek  and Month as possible factors. 

```{r include=TRUE}
df$Origin <- as.factor(df$Origin)
class(df$Origin)
```

---
#Categorical Variables as Bar Plots

```{r}
plot(df$Origin)
```

---


```{r}
summary(df)
```

add describe() here


---
#Univariate analysis

In univariate analysis, the output is a single variable and all data collected is for it. There is no cause-and-effect relationship at all

## Charts to use

- Categorical ( Bar , Pie Charts)
- Quantitative (Histogram, BoxPlot, DensityPlot)

---
```{r echo=FALSE}
df$DayOfWeek <- as.factor(df$DayOfWeek)
df$Month <- as.factor(df$Month)
df$UniqueCarrier <- as.factor(df$UniqueCarrier)
```

#Categorical Variables

.pull-left[
```{r}
plot(df$Month)

```
]
.pull-right[
```{r}
 plot(df$DayOfWeek)
```

]
---
#Categorical Variables
```{r}
 plot(df$UniqueCarrier)
```


---
#Quantitative Variables



.pull-left[
```{r}
hist(df$DepDelay)

```
]
.pull-right[
```{r}
 hist(df$ArrDelay)
```
  

]
---
#BiVariate Analysis

Bivariate analysis refers to the analysis of two variables to determine **relationships** between them. One variable here is dependent while the other is independent. 

###Types of bivariate data analysis

- **Numerical and Numerical** – In this type, both the variables of bivariate data, independent and dependent, - are having numerical values.

- **Categorical and Categorical** – When both the variables are categorical.

- **Numerical and Categorical** – When one variable is numerical and one is categorical.

---

Numerical and Numerical
```{r }
# Get only flights from American Airlines
AA <- subset(hflights, UniqueCarrier == "AA" )
plot(x= AA$ArrDelay , y= AA$DepDelay,  xlab = "Arrival Delay",  ylab = "Departure Delay", col = "dodgerblue1")
```
---
Numerical and Categorical
```{r}
plot(AA$ArrDelay ~ AA$DayOfWeek , xlab = "Day of Week",  ylab = "Arrival Delay", col = "dodgerblue1")
```
---
Categorical and Categorical

```{r}
plot(as.factor(df$DayOfWeek) ~ as.factor(df$Origin))

```
---
class: inverse, center, middle

#Next : Data Manipulation



